\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-03-04 paper.tex
fix error in lambda_0 update
Update the complete log likelihood formula
Update the expected Lambda and Monte-Carlo sample set

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that for ease of notations, we include $\lambda_0$ in $\theta$ although it is a function.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\} \]
and
\begin{equation}
  \label{eq:b_density}
  f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.
\end{equation}

@@

In the case $g_i(b_i, G_i) = \tilde g_i(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g_i(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:Lambda_ik-def}
  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
\end{equation}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice. Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context. We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result. Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S^{(w)}$ used for the approximation.

@@

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{Construction of the samples in $S^{(w)}$}
\label{alg:construct-set-S}
\KwInput{$S^{(w)} = \varnothing$}
\KwOutput{$S^{(w)}$ filled with $2N$ samples}
Compute $C^{(w)} \in \R^{r \times r}$ such that $C^{(w)}{C^{(w)}}^\top = D^{(w)}$ \tcp{Cholesky decomposition}
\For{$\breve{n} = 1,\ldots,N$}    
{ 
  Sample $\Omega_{\breve{n}} \sim \cN(0, I_r)$ \\
  Compute $\breve{b}_{\breve{n}} = C^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
  Update $S^{(w)} \leftarrow S^{(w)} \cup \{\breve{b}_{\breve{n}}, -\breve{b}_{\breve{n}} \}$
}
\textbf{Return:} {$S^{(w)}$}
\end{algorithm}

@@

\noindent The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
\begin{equation}
  \label{eq:def-hat-E-g-tilde}
  \hat \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:hat_Lambda_ik}
  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
\end{equation}

>>>>

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that for ease of notations, we include $\lambda_0$ in $\theta$ although it is a function.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(y_i | G_i = k)\int_{\R^r} f(t_i, \delta_i| b_i, G_i = k ; \theta) f(b_i | y_i, G_i = k; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\} \]
and
\begin{align}
  \label{eq:b_density}
  b_i | y_i, G_i=k; \theta &\sim \cN\Big((V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}V_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}\Big) \nonumber \\
  y_i | G_i = k; \theta &\sim \cN\Big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\Big)
\end{align}

@@

In the case $g_i(b_i, G_i) = \tilde g_i(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g_i(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:Lambda_ik-def}
  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i,
\end{equation}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice. Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context. We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result. Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S^{(w)}$ used for the approximation.

@@

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{Construction of the samples in $S_i^{(w)}$}
\label{alg:construct-set-S}
\KwInput{$S_i^{(w)} = \varnothing$}
\KwOutput{$S_i^{(w)}$ filled with $2N$ samples}
Compute $C_i^{(w)} \in \R^{r \times r}$ such that $C_i^{(w)}{C_i^{(w)}}^\top = (V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1})^{-1}$ \tcp{Cholesky decomposition}
\For{$\breve{n} = 1,\ldots,N$}    
{ 
  Sample $\Omega_{\breve{n}} \sim \cN(0, I_r)$ \\
  Compute $\breve{b}_{\breve{n}} = C_i^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
  Update $S_i^{(w)} \leftarrow S_i^{(w)} \cup \{\breve{b}_{\breve{n}}, -\breve{b}_{\breve{n}} \}$
}
\textbf{Return:} {$S_i^{(w)}$}
\end{algorithm}

@@

\noindent The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
\begin{equation}
  \label{eq:def-hat-E-g-tilde}
  \hat \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:hat_Lambda_ik}
  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_i^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
\end{equation}
